Fusing Filters with Integer Linear Programming

### 2014-05-04 2014betterfusionforfilters/section/03-SizeInference.tex
size inf: constraints: change 'satisfiable' to 'valid'

The most important constraint is equality on two sizes.
Context holes are used to simplify later definitions, and the notation $\fillhole{c}{d}$ is used to denote replacing all holes in $c$ with $d$.
Constraints may be joined together with $c \wedge d$, which requires both $c$ and $d$ to be satisfiable.
Forall quantifiers mean the constraint must be satisfiable for every size.
Existential quantifiers simply state that there must exist a size that satisfies the constraint.
\begin{tabbing}
MMMMM   \= MM \= MMMMMMMMM \= \kill
$C$     \> @::=@ \> $true$              \> (trivially true)             \\
%       \> $~|$  \> $\hole$             \> (constraint hole)            \\
        \> $~|$  \> $size = size$       \> (equality constraint)        \\
        \> $~|$  \> $C \wedge C$        \> (conjunction)                \\
        \> $~|$  \> $\forall k_n.\ C$   \> (forall quantification)      \\
        \> $~|$  \> $\exists k_n.\ C$   \> (existential quantification)
\end{tabbing}

@@

% -----------------------------------------------------------------------------
\subsubsection{Examples}
As an example of constraint generation, here is the @normalize2@ program from earlier.
\begin{tabbing}
@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
@normalize2 us@                     \> $\exists k_{us}.$      \\
@ = let sum1 = fold   (+) 0 us@     \>                      \\
@       gts  = filter (>0)  us@     \> $\forall k_{gts}.$ \\
@       sum2 = fold   (+) 0 gts@    \> \\
@       nor1 = map  (/sum1) us@     \> $\exists k_{nor1}.\ k_{nor1} = k_{us} \wedge$ \\
@       nor2 = map  (/sum2) us@     \> $\exists k_{nor2}.\ k_{nor2} = k_{us} \wedge$ \\
@   in (nor1, nor2)@                \> $true$ \\
\end{tabbing}
This constraint is satisfiable, with the equivalence classes being:
\newcommand{\eqclasses}[1]{
    \begin{tabbing}
        MM \= M \= \kill
        #1
    \end{tabbing}}
\newcommand{\eqclass}[2]{$#1$ \> $\in$ \> $\{#2\}$ \\}
\eqclasses{
    \eqclass{k_{us}}{k_{us}, k_{nor1}, k_{nor2}}
    \eqclass{k_{gts}}{k_{gts}}
}

@@

This constraint is satisfiable, with the equivalence classes being:
\eqclasses{
    \eqclass{k_{xs}}    {k_{xs}}
    \eqclass{k_{ys1}}   {k_{ys1}}
    \eqclass{k_{ys2}}   {k_{ys2}}
}

This example is disallowed, as it would require a runtime check on the size of the @flt@ array.
\begin{tabbing}
@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
@bad1 vs@                           \> $\exists k_{vs}.$ \\
@ = let flt   = filter p vs@        \> $\forall k_{flt}.$ \\
@       wrong = map2   f flt vs@    \> $\exists k_{wrong}.\ k_{wrong} = k_{flt}$ \\
                                    \> $\wedge k_{wrong} = k_{vs} \wedge$ \\
@   in  wrong@                      \> $true$
\end{tabbing}
This constraint is unsatisfiable, as by transitivity it can be simplified to $\exists k_{vs}.\ \forall k_{flt}.\ k_{vs} = k_{flt}$.
Obviously, there exists no sole size such that all sizes are equal to it.
As a result, no equivalence classes are generated for this example, and no fusion is performed.

Similarly, this example is not allowed, as it would require a runtime check to determine that the output of two filters @flt@ and @flt2@ are the same size.
\begin{tabbing}
@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
@bad2 vs@                           \> $\exists k_{vs}.$ \\
@ = let flt  = filter p  vs@        \> $\forall k_{flt}.$ \\
@       flt2 = filter p' vs@        \> $\forall k_{flt2}.$ \\
@       mix  = map2   f  flt flt2@  \> $\exists k_{mix}.\ k_{mix} = k_{flt}$ \\
                                    \> $\wedge k_{mix} = k_{flt2} \wedge$ \\
@   in  mix@                        \> $true$                               \\
\end{tabbing}
Again, by transitivity of equality this constraint can be simplified to $\forall k_{flt}.\ \forall k_{flt2}.\ k_{flt} = k_{flt2}$.
It is easy to see that this constraint is unsatisfiable.

@@

Once the constraints are solved, known to be satisfiable, and sorted into equivalence classes, each combinator is assigned a size.
Note that for a filter, the size of the output array $k_o$ is some existential that is less than or equal to $k_n$, but the actual loop size of the \emph{combinator} is equal to $k_n$.
This is because, in order to produce the filtered output, all elements of the input $n$ must be considered.

>>>>

The most important constraint is equality on two sizes.
Context holes are used to simplify later definitions, and the notation $\fillhole{c}{d}$ is used to denote replacing all holes in $c$ with $d$.
Constraints may be joined together with $c \wedge d$, which requires both $c$ and $d$ to be valid.
Forall quantifiers mean the constraint must be satisfiable for every size.
Existential quantifiers simply state that there must exist a size that satisfies the constraint.
\begin{tabbing}
MMMMM   \= MM \= MMMMMMMMM \= \kill
$C$     \> @::=@ \> $true$              \> (trivially true)             \\
%       \> $~|$  \> $\hole$             \> (constraint hole)            \\
        \> $~|$  \> $size = size$       \> (equality constraint)        \\
        \> $~|$  \> $C \wedge C$        \> (conjunction)                \\
        \> $~|$  \> $\forall k_n.\ C$   \> (forall quantification)      \\
        \> $~|$  \> $\exists k_n.\ C$   \> (existential quantification)
\end{tabbing}

@@

% -----------------------------------------------------------------------------
\subsubsection{Examples}
As an example of constraint generation, here is the @normalize2@ program from earlier.
\begin{tabbing}
@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
@normalize2 us@                     \> $\exists k_{us}.$      \\
@ = let sum1 = fold   (+) 0 us@     \>                      \\
@       gts  = filter (>0)  us@     \> $\forall k_{gts}.$ \\
@       sum2 = fold   (+) 0 gts@    \> \\
@       nor1 = map  (/sum1) us@     \> $\exists k_{nor1}.\ k_{nor1} = k_{us} \wedge$ \\
@       nor2 = map  (/sum2) us@     \> $\exists k_{nor2}.\ k_{nor2} = k_{us} \wedge$ \\
@   in (nor1, nor2)@                \> $true$ \\
\end{tabbing}
This constraint is valid, with the equivalence classes being:
\newcommand{\eqclasses}[1]{
    \begin{tabbing}
        MM \= M \= \kill
        #1
    \end{tabbing}}
\newcommand{\eqclass}[2]{$#1$ \> $\in$ \> $\{#2\}$ \\}
\eqclasses{
    \eqclass{k_{us}}{k_{us}, k_{nor1}, k_{nor2}}
    \eqclass{k_{gts}}{k_{gts}}
}

@@

This constraint is valid, with the equivalence classes being:
\eqclasses{
    \eqclass{k_{xs}}    {k_{xs}}
    \eqclass{k_{ys1}}   {k_{ys1}}
    \eqclass{k_{ys2}}   {k_{ys2}}
}

This example is disallowed, as it would require a runtime check on the size of the @flt@ array.
\begin{tabbing}
@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
@bad1 vs@                           \> $\exists k_{vs}.$ \\
@ = let flt   = filter p vs@        \> $\forall k_{flt}.$ \\
@       wrong = map2   f flt vs@    \> $\exists k_{wrong}.\ k_{wrong} = k_{flt}$ \\
                                    \> $\wedge k_{wrong} = k_{vs} \wedge$ \\
@   in  wrong@                      \> $true$
\end{tabbing}
This constraint is invalid, as by transitivity it can be simplified to $\exists k_{vs}.\ \forall k_{flt}.\ k_{vs} = k_{flt}$.
Obviously, there exists no sole size such that all sizes are equal to it.
As a result, no equivalence classes are generated for this example, and no fusion is performed.

Similarly, this example is not allowed, as it would require a runtime check to determine that the output of two filters @flt@ and @flt2@ are the same size.
\begin{tabbing}
@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
@bad2 vs@                           \> $\exists k_{vs}.$ \\
@ = let flt  = filter p  vs@        \> $\forall k_{flt}.$ \\
@       flt2 = filter p' vs@        \> $\forall k_{flt2}.$ \\
@       mix  = map2   f  flt flt2@  \> $\exists k_{mix}.\ k_{mix} = k_{flt}$ \\
                                    \> $\wedge k_{mix} = k_{flt2} \wedge$ \\
@   in  mix@                        \> $true$                               \\
\end{tabbing}
Again, by transitivity of equality this constraint can be simplified to $\forall k_{flt}.\ \forall k_{flt2}.\ k_{flt} = k_{flt2}$.
It is easy to see that this constraint is invalid.

@@

Once the constraints are solved, known to be valid, and sorted into equivalence classes, each combinator is assigned a size.
Note that for a filter, the size of the output array $k_o$ is some existential that is less than or equal to $k_n$, but the actual loop size of the \emph{combinator} is equal to $k_n$.
This is because, in order to produce the filtered output, all elements of the input $n$ must be considered.
